---
title: "Capstone Project"
author: "Loan Portfolio Prediction Team"
date: "2023-10-04"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE,  warning=FALSE}
library(readr)
library("tidyverse")
library("skimr")
library("readxl") # used to read excel files
library("dplyr") # used for data munging 
library("FNN") # used for knn regression (knn.reg function)
library("caret") # used for various predictive models
library("class") # for using confusion matrix function
library("rpart.plot") # used to plot decision tree
library("rpart")  # used for Regression tree
library("glmnet") # used for Lasso and Ridge regression
library('NeuralNetTools') # used to plot Neural Networks
library("PRROC") # top plot ROC curve
library("ROCR") # top plot lift curve
library("C50")
library("kableExtra")
library("formattable")
library("reticulate")
library(tm) #For text mining functionality
library(SnowballC) #For collapsing words to common roots
library(wordcloud) #For creating word cloud visualization
library(RColorBrewer) #For creating colorful graphs using pre-defined palettes
library(caTools) #For splitting the data  
library(ranger) # used to random forest
library(randomForest) # used to random forest
library(quanteda) #to get ngrams without using RWeka
library(corrplot)
library("FNN") # used for knn regression (knn.reg function)
library("class") # for using confusion matrix function
library("glmnet") # used for Lasso and Ridge regression
library('NeuralNetTools') # used to plot Neural Networks
library("tidyverse")
library("skimr")
library(randomForest)
library(readxl)
library(rpart)
library(DMwR)
library(ROSE)

```


#cleaning the train data set
```{r}
# Load the necessary library
#library(readxl)

# Specify the file path and sheet name
file_path <- "Project Data observed in 2 years -1004.xlsx"
sheet_name <- "data 25993"

# Read the specified sheet and let read_excel infer the column types
whole_data <- read_excel(path = file_path, sheet = sheet_name)

# Exclude the first column, item column
whole_data <- select(whole_data, -1)

# Convert each column to factor explicitly
whole_data[] <- lapply(whole_data, as.factor)

# Check the classes of each column to confirm
attribute_classes <- sapply(whole_data, class)
print(attribute_classes)

```

```{r}
# Count the number of NA values in each column

na_count <- sapply(whole_data, function(x) sum(is.na(x)))

# Print the count of NA values for each column
print(na_count)

# Apply the table function to each column of the data frame
table_list <- lapply(whole_data, table)

# Print the table of each column
for (i in seq_along(table_list)) {
  cat("Table for column:", names(table_list)[i], "\n")
  print(table_list[[i]])
  cat("\n")
}

table_list

```


```{r}
library(dplyr)

whole_data <- whole_data %>%
  mutate(target_variable = ifelse(`survial year` == "Normal case", 0, 1))

# Check the classes of each column to confirm
attribute_classes <- sapply(whole_data, class)
print(attribute_classes)
table(whole_data$target_variable)

```
```{r}
# Data pre-processing
# Select only the change variables and the response variable
predictor_columns  <- c(
  "G_total_limit",
  "G_risk_grade",
  "G_trade_bureau_cnt",
  "G_period_employment",
  "G_house_status",
  "G_net_income",
  "G_race",
  "G_enquiry_cnt"
)

model_data <- whole_data %>%
  select(predictor_columns , target_variable ) %>%
  na.omit()

# Check the classes of each column to confirm
attribute_classes <- sapply(model_data, class)
print(attribute_classes)

table(model_data$target_variable)
```

```{r}
# Split the whole data into training and test sets (80% train, 20% test)

# Load caret library
library(caret)

# Check distribution of target_variable
table(model_data$target_variable)

# If distribution seems okay, manually split the data
set.seed(123)

# Create a randomized index for splitting
index <- sample(seq_len(nrow(model_data)), size = 0.80 * nrow(model_data))

train_data <- model_data[index, ]
test_data <- model_data[-index, ]

table(train_data$target_variable)
table(test_data$target_variable)
```

#Task2.You will use the predictive modeling process we learnt in the class to land on a final binary classification model to use. 
#a. Start by building a model using only the previous time period’s change variables

# 1. Logistic Regression Model building, threshold=0.5
```{r}

logistic_regression_model <- glm(target_variable ~ ., data = train_data, family = "binomial")

#summary(logistic_regression_model)

# Make predictions on the test set
logistic_regression_prob_predictions <- predict(logistic_regression_model, newdata = test_data, type = "response")


# Use the optimal threshold to convert probabilities to class labels
logistic_regression_predictions <- ifelse(logistic_regression_prob_predictions >= 0.5, 1, 0)

test_data$target_variable <- as.factor(test_data$target_variable)

# Convert the numeric predictions to factor
logistic_regression_predictions <- as.factor(logistic_regression_predictions)

#levels(test_data$target_variable)
#print(test_data$target_variable)

levels(logistic_regression_predictions) <- levels(test_data$target_variable)

# Model evaluation 
logistic_regression_confusion_matrix <- confusionMatrix(logistic_regression_predictions, test_data$target_variable)
print(logistic_regression_confusion_matrix)

# Extract and print overall statistics and class-wise statistics
overall_stats_logistic_regression <- logistic_regression_confusion_matrix[["overall"]]
classwise_stats_logistic_regression <- logistic_regression_confusion_matrix[["byClass"]]

cat("Accuracy for Logistic Regression:", round(overall_stats_logistic_regression[["Accuracy"]], 3), "\n")
cat("Precision for Logistic Regression:", round(classwise_stats_logistic_regression[["Precision"]], 3), "\n")
cat("Recall for Logistic Regression:", round(classwise_stats_logistic_regression[["Recall"]], 3), "\n")

# Calculate F1 score
precision_logistic_regression <- classwise_stats_logistic_regression[["Precision"]]
recall_logistic_regression <- classwise_stats_logistic_regression[["Recall"]]
f1_score_logistic_regression <- 2 * (precision_logistic_regression * recall_logistic_regression) / (precision_logistic_regression + recall_logistic_regression)

cat("F1 Score for Logistic Regression:", round(f1_score_logistic_regression, 3), "\n")

table(train_data$G_net_income)
table(test_data$G_net_income)
```

```{r}

# Here is the code I wrote (Mingi)


library(dplyr)
library(caret)

# Apply transformation to the train data set
train_data <- train_data %>%
  mutate(G_net_income = case_when(
    row_number() == 1 & G_net_income == "N < 1,000" ~ "N < 1,000", # Leave the first observation unchanged
    G_net_income == "N < 1,000" ~ "1,000 =< N <1.500", # Change the rest of the observations
    G_net_income == "1,000 =< N <1.500" ~ "1,500 =< N < 2,000",
    G_net_income == "1,500 =< N < 2,000" ~ "2,000 =< N",
    TRUE ~ G_net_income # Default case
  ))

table(train_data$G_net_income) # Check the changes in train_data$G_net_income

test_data <- test_data %>%
  mutate(G_net_income = case_when(
    row_number() == 1 & G_net_income == "N < 1,000" ~ "N < 1,000",
    G_net_income == "N < 1,000" ~ "1,000 =< N <1.500",
    G_net_income == "1,000 =< N <1.500" ~ "1,500 =< N < 2,000",
    G_net_income == "1,500 =< N < 2,000" ~ "2,000 =< N",
    TRUE ~ G_net_income # Default case
  ))
table(test_data$G_net_income) # Check the changes in test_data$G_net_income

# Train the logistic regression model with modified data
logistic_regression_model_counterfactual <- glm(target_variable ~ ., data = train_data, family = "binomial")

# Make predictions on the test set with the counterfactual model
logistic_regression_prob_predictions_counterfactual <- predict(logistic_regression_model_counterfactual, newdata = test_data, type = "response")

# Use the optimal threshold to convert probabilities to class labels
logistic_regression_predictions_counterfactual <- ifelse(logistic_regression_prob_predictions_counterfactual >= 0.5, 1, 0)

test_data$target_variable <- as.factor(test_data$target_variable)

# Convert the numeric predictions to factor
logistic_regression_predictions_counterfactual <- as.factor(logistic_regression_predictions_counterfactual)

levels(logistic_regression_predictions_counterfactual) <- levels(test_data$target_variable)

# Model evaluation for the counterfactual model
logistic_regression_confusion_matrix_counterfactual <- confusionMatrix(logistic_regression_predictions_counterfactual, test_data$target_variable)
print(logistic_regression_confusion_matrix_counterfactual)

# Extract and print overall statistics and class-wise statistics for the counterfactual model
overall_stats_logistic_regression_counterfactual <- logistic_regression_confusion_matrix_counterfactual[["overall"]]
classwise_stats_logistic_regression_counterfactual <- logistic_regression_confusion_matrix_counterfactual[["byClass"]]

cat("Accuracy for Logistic Regression (Counterfactual):", round(overall_stats_logistic_regression_counterfactual[["Accuracy"]], 3), "\n")
cat("Precision for Logistic Regression (Counterfactual):", round(classwise_stats_logistic_regression_counterfactual[["Precision"]], 3), "\n")
cat("Recall for Logistic Regression (Counterfactual):", round(classwise_stats_logistic_regression_counterfactual[["Recall"]], 3), "\n")

# Calculate F1 score for the counterfactual model
precision_logistic_regression_counterfactual <- classwise_stats_logistic_regression_counterfactual[["Precision"]]
recall_logistic_regression_counterfactual <- classwise_stats_logistic_regression_counterfactual[["Recall"]]
f1_score_logistic_regression_counterfactual <- 2 * (precision_logistic_regression_counterfactual * recall_logistic_regression_counterfactual) / (precision_logistic_regression_counterfactual + recall_logistic_regression_counterfactual)

cat("F1 Score for Logistic Regression (Counterfactual):", round(f1_score_logistic_regression_counterfactual, 3), "\n")



```



```{r}



# Clean and preprocess the train data set
# Assuming you have already loaded libraries and read the data

# Modify the "G_net_income" variable
library(dplyr)

train_data <- train_data %>%
  mutate(G_net_income = case_when(
    row_number() == 1 & G_net_income == "N < 1,000" ~ "N < 1,000", # Leave the first observation unchanged
    G_net_income == "N < 1,000" ~ "1,000 =< N < 1,500", # Change the rest of the observations
    G_net_income == "1,000 =< N < 1,500" ~ "1,500 =< N < 2,000",
    G_net_income == "1,500 =< N < 2,000" ~ "2,000 =< N",
    TRUE ~ G_net_income # Leaves other values unchanged
  ))



# Train the logistic regression model with the modified data
logistic_regression_model_counterfactual <- glm(target_variable ~ ., data = train_data, family = "binomial")

# Update factor levels in test_data to match training data
test_data$G_net_income <- factor(test_data$G_net_income, levels = levels(train_data$G_net_income))

# Make predictions on the test set with the modified model
logistic_regression_prob_predictions_counterfactual <- predict(logistic_regression_model_counterfactual, newdata = test_data, type = "response")


# Use the optimal threshold to convert probabilities to class labels
logistic_regression_predictions_counterfactual <- ifelse(logistic_regression_prob_predictions_counterfactual >= 0.5, 1, 0)

# Convert the numeric predictions to factor
logistic_regression_predictions_counterfactual <- as.factor(logistic_regression_predictions_counterfactual)

# Set the levels of the predictions to match the levels of the actual target variable
levels(logistic_regression_predictions_counterfactual) <- levels(test_data$target_variable)

# Evaluate the performance of the counterfactual model
logistic_regression_confusion_matrix_counterfactual <- confusionMatrix(logistic_regression_predictions_counterfactual, test_data$target_variable)
print(logistic_regression_confusion_matrix_counterfactual)

# Extract and print overall statistics and class-wise statistics
overall_stats_logistic_regression_counterfactual <- logistic_regression_confusion_matrix_counterfactual[["overall"]]
classwise_stats_logistic_regression_counterfactual <- logistic_regression_confusion_matrix_counterfactual[["byClass"]]

cat("Accuracy for Logistic Regression (Counterfactual):", round(overall_stats_logistic_regression_counterfactual[["Accuracy"]], 3), "\n")
cat("Precision for Logistic Regression (Counterfactual):", round(classwise_stats_logistic_regression_counterfactual[["Precision"]], 3), "\n")
cat("Recall for Logistic Regression (Counterfactual):", round(classwise_stats_logistic_regression_counterfactual[["Recall"]], 3), "\n")

# Calculate F1 score for the counterfactual model
precision_logistic_regression_counterfactual <- classwise_stats_logistic_regression_counterfactual[["Precision"]]
recall_logistic_regression_counterfactual <- classwise_stats_logistic_regression_counterfactual[["Recall"]]
f1_score_logistic_regression_counterfactual <- 2 * (precision_logistic_regression_counterfactual * recall_logistic_regression_counterfactual) / (precision_logistic_regression_counterfactual + recall_logistic_regression_counterfactual)

cat("F1 Score for Logistic Regression (Counterfactual):", round(f1_score_logistic_regression_counterfactual, 3), "\n")

```










# 2. Decision Tree Model building
```{r}
# Train the decision tree model
tree_model <- rpart(target_variable ~ ., data = train_data, method = "class")

# Make predictions on the test set
tree_prob_predictions <- predict(tree_model, newdata = test_data, type = "prob")

# The predict function will return a matrix. We'll extract probabilities for the positive class
positive_probs <- tree_prob_predictions[,2]

# Use a threshold to convert probabilities to class labels
tree_predictions <- ifelse(positive_probs >= 0.5, 1, 0)

# Convert the numeric predictions to factor
tree_predictions <- as.factor(tree_predictions)
test_data$target_variable <- as.factor(test_data$target_variable)

# Match the levels of predictions to the actual target variable
levels(tree_predictions) <- levels(test_data$target_variable)

# Model evaluation 
tree_confusion_matrix <- confusionMatrix(tree_predictions, test_data$target_variable)
print(tree_confusion_matrix)

# Extract and print overall statistics and class-wise statistics
overall_stats_tree <- tree_confusion_matrix[["overall"]]
classwise_stats_tree <- tree_confusion_matrix[["byClass"]]

cat("Accuracy for Decision Tree:", round(overall_stats_tree[["Accuracy"]], 3), "\n")
cat("Precision for Decision Tree:", round(classwise_stats_tree[["Precision"]], 3), "\n")
cat("Recall for Decision Tree:", round(classwise_stats_tree[["Recall"]], 3), "\n")

# Calculate F1 score for Decision Tree
precision_tree <- classwise_stats_tree[["Precision"]]
recall_tree <- classwise_stats_tree[["Recall"]]
f1_score_tree <- 2 * (precision_tree * recall_tree) / (precision_tree + recall_tree)

cat("F1 Score for Decision Tree:", round(f1_score_tree, 3), "\n")

```

# 3. Random Forest Model building
```{r}
# Load required libraries
library(randomForest)
library(caret)

# Ensure that the target variable in the training data is a factor
train_data$target_variable <- as.factor(train_data$target_variable)

# Train a Random Forest model for classification
rf_model <- randomForest(target_variable ~ ., data=train_data, ntree=100, mtry=3, importance=TRUE)

# Make predictions on the test set
rf_predictions <- predict(rf_model, newdata=test_data, type="response")

# Ensure that the predictions and test_data target variable are both factors
rf_predictions <- as.factor(rf_predictions)
test_data$target_variable <- as.factor(test_data$target_variable)

# Evaluate model predictions using a confusion matrix
rf_confusion_matrix <- confusionMatrix(rf_predictions, test_data$target_variable)
print(rf_confusion_matrix)

# Extract and print overall statistics and class-wise statistics
overall_stats_rf <- rf_confusion_matrix[["overall"]]
classwise_stats_rf <- rf_confusion_matrix[["byClass"]]

cat("Accuracy for Random Forest:", round(overall_stats_rf[["Accuracy"]], 3), "\n")
cat("Precision for Random Forest:", round(classwise_stats_rf[["Precision"]], 3), "\n")
cat("Recall for Random Forest:", round(classwise_stats_rf[["Recall"]], 3), "\n")

# Calculate F1 score
precision_rf <- classwise_stats_rf[["Precision"]]
recall_rf <- classwise_stats_rf[["Recall"]]
f1_score_rf <- 2 * (precision_rf * recall_rf) / (precision_rf + recall_rf)
cat("F1 Score for Random Forest:", round(f1_score_rf, 3), "\n")

```
# 4. KNN Model building
```{r}

# Load necessary libraries
library(class)
library(caret)

# Convert target variable to factor for consistency in predictions and evaluation
train_data$target_variable <- as.factor(train_data$target_variable)
test_data$target_variable <- as.factor(test_data$target_variable)

#attribute_classes_train_knn <- sapply(train_data, class)
#print(attribute_classes_train_knn)

#attribute_classes_test_knn <- sapply(test_data, class)
#print(attribute_classes_test_knn)

# Exclude the target_variable for encoding and split it off into its own vector
train_labels <- train_data$target_variable
test_labels <- test_data$target_variable

train_matrix <- model.matrix(~. - 1 - target_variable, data = train_data)
test_matrix <- model.matrix(~. - 1 - target_variable, data = test_data)

# Run k-NN
knn_predictions <- knn(train_matrix, test_matrix, cl=train_labels, k=3)

library(caret)

# Ensure the predictions and test_labels are both factors
knn_predictions <- as.factor(knn_predictions)
test_labels <- as.factor(test_labels)

# Evaluate model predictions using a confusion matrix
knn_confusion_matrix <- confusionMatrix(knn_predictions, test_labels)
print(knn_confusion_matrix)

# Extract and print overall statistics and class-wise statistics
overall_stats_knn <- knn_confusion_matrix[["overall"]]
classwise_stats_knn <- knn_confusion_matrix[["byClass"]]

cat("Accuracy for k-NN:", round(overall_stats_knn[["Accuracy"]], 3), "\n")
cat("Precision for k-NN:", round(classwise_stats_knn[["Precision"]], 3), "\n")
cat("Recall for k-NN:", round(classwise_stats_knn[["Recall"]], 3), "\n")

# Calculate F1 score
precision_knn <- classwise_stats_knn[["Precision"]]
recall_knn <- classwise_stats_knn[["Recall"]]
f1_score_knn <- 2 * (precision_knn * recall_knn) / (precision_knn + recall_knn)
cat("F1 Score for k-NN:", round(f1_score_knn, 3), "\n")

```


#Performace comparison each model
```{r}
library(caret)
library(pROC) # for AUC calculation

# Create an empty matrix to store model performances
model_performance <- matrix(0, nrow = 4, ncol = 5)
rownames(model_performance) <- c("Logistic Regression", "Decision Tree", "Random Forest", "kNN")
colnames(model_performance) <- c("Accuracy", "Precision", "Recall", "F1 Score", "AUC")

# 1. Logistic Regression
overall_stats_logistic_regression <- logistic_regression_confusion_matrix[["overall"]]
classwise_stats_logistic_regression <- logistic_regression_confusion_matrix[["byClass"]]

model_performance["Logistic Regression", ] <- c(
  overall_stats_logistic_regression[["Accuracy"]],
  classwise_stats_logistic_regression[["Precision"]],
  classwise_stats_logistic_regression[["Recall"]],
  2 * (classwise_stats_logistic_regression[["Precision"]] * classwise_stats_logistic_regression[["Recall"]]) /
    (classwise_stats_logistic_regression[["Precision"]] + classwise_stats_logistic_regression[["Recall"]]),
  auc(response = as.numeric(test_data$target_variable), predictor = logistic_regression_prob_predictions)
)

# 2. Decision Tree
# You'll need to replace the decisionTree_confusion_matrix with the actual confusion matrix for your Decision Tree Model
overall_stats_tree <- tree_confusion_matrix[["overall"]]
classwise_stats_tree <- tree_confusion_matrix[["byClass"]]


model_performance["Decision Tree", ] <- c(
  overall_stats_tree[["Accuracy"]],
  classwise_stats_tree[["Precision"]],
  classwise_stats_tree[["Recall"]],
  classwise_stats_tree[["F1"]],
  roc(as.numeric(as.factor(test_data$target_variable)), as.numeric(tree_predictions))$auc
)

# 3. Random Forest
overall_stats_rf <- rf_confusion_matrix[["overall"]]
classwise_stats_rf <- rf_confusion_matrix[["byClass"]]


model_performance["Random Forest", ] <- c(
  overall_stats_rf[["Accuracy"]],
  classwise_stats_rf[["Precision"]],
  classwise_stats_rf[["Recall"]],
  f1_score_rf,
  auc(response = as.numeric(test_data$target_variable), predictor = as.numeric(rf_predictions))
)

# 4. kNN
overall_stats_knn <- knn_confusion_matrix[["overall"]]
classwise_stats_knn <- knn_confusion_matrix[["byClass"]]


model_performance["kNN", ] <- c(
  overall_stats_knn[["Accuracy"]],
  classwise_stats_knn[["Precision"]],
  classwise_stats_knn[["Recall"]],
  2 * (classwise_stats_knn[["Precision"]] * classwise_stats_knn[["Recall"]]) / 
    (classwise_stats_knn[["Precision"]] + classwise_stats_knn[["Recall"]]),
  auc(response = as.numeric(test_data$target_variable), predictor = as.numeric(knn_predictions))
)

# Print the model performance matrix
print(model_performance)
```

