---
title: "Capstone Project"
author: "Loan Portfolio Prediction Team"
date: "2023-10-16"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE,  warning=FALSE}
library(readr)
library("tidyverse")
library("skimr")
library("readxl") # used to read excel files
library("dplyr") # used for data munging 
library("FNN") # used for knn regression (knn.reg function)
library("caret") # used for various predictive models
library("class") # for using confusion matrix function
library("rpart.plot") # used to plot decision tree
library("rpart")  # used for Regression tree
library("glmnet") # used for Lasso and Ridge regression
library('NeuralNetTools') # used to plot Neural Networks
library("PRROC") # top plot ROC curve
library("ROCR") # top plot lift curve
library("C50")
library("kableExtra")
library("formattable")
library("reticulate")
library(tm) #For text mining functionality
library(SnowballC) #For collapsing words to common roots
library(wordcloud) #For creating word cloud visualization
library(RColorBrewer) #For creating colorful graphs using pre-defined palettes
library(caTools) #For splitting the data  
library(ranger) # used to random forest
library(randomForest) # used to random forest
library(quanteda) #to get ngrams without using RWeka
library(corrplot)
library("FNN") # used for knn regression (knn.reg function)
library("class") # for using confusion matrix function
library("glmnet") # used for Lasso and Ridge regression
library('NeuralNetTools') # used to plot Neural Networks
library("tidyverse")
library("skimr")
library(randomForest)
library(readxl)
library(rpart)
library(DMwR)
library(ROSE)
library(DMwR)
library(pROC)
library(ROSE)  # for dataset balancing

```


#cleaning the train data set
```{r}
# Load the necessary library
#library(readxl)

# Specify the file path and sheet name
file_path <- "C:/Users/MY00002/Documents/NYU/Course/Capstone/Data/data/Project Data observed in 2 years -1016.xlsx"
sheet_name <- "data 56355(1st)"

# Read the specified sheet and let read_excel infer the column types
whole_data <- read_excel(path = file_path, sheet = sheet_name)

# Exclude the first column, item column
whole_data <- select(whole_data, -1)

# Convert each column to factor explicitly
whole_data[] <- lapply(whole_data, as.factor)

# Check the classes of each column to confirm
attribute_classes <- sapply(whole_data, class)
print(attribute_classes)

```

```{r}
# Count the number of NA values in each column

na_count <- sapply(whole_data, function(x) sum(is.na(x)))

# Print the count of NA values for each column
print(na_count)

# Apply the table function to each column of the data frame
table_list <- lapply(whole_data, table)

# Print the table of each column
for (i in seq_along(table_list)) {
  cat("Table for column:", names(table_list)[i], "\n")
  print(table_list[[i]])
  cat("\n")
}


```


```{r}
library(dplyr)

whole_data <- whole_data %>%
  mutate(target_variable = ifelse(`survival year` == "Normal case", 0, 1))

# Check the classes of each column to confirm
attribute_classes <- sapply(whole_data, class)
print(attribute_classes)
table(whole_data$target_variable)

```
```{r}
# Data pre-processing
# Select only the change variables and the response variable
predictor_columns  <- c(
  "G_total_limit",
  "G_risk_grade",
  "G_trade_bureau_cnt",
  "G_period_employment",
  "G_house_status",
  "G_net_income",
  "G_race",
  "G_enquiry_cnt"
)

model_data <- whole_data %>%
  select(predictor_columns , target_variable ) %>%
  na.omit()

# Check the classes of each column to confirm
attribute_classes <- sapply(model_data, class)
print(attribute_classes)

table(model_data$target_variable)
```

```{r}
# Split the whole data into training and test sets (80% train, 20% test)

# Load caret library
library(caret)

# Check distribution of target_variable
table(model_data$target_variable)

# If distribution seems okay, manually split the data
set.seed(12345)

# Create a randomized index for splitting
index <- sample(seq_len(nrow(model_data)), size = 0.80 * nrow(model_data))

train_data <- model_data[index, ]
test_data <- model_data[-index, ]

table(train_data$target_variable)
table(test_data$target_variable)
```

#Task2.You will use the predictive modeling process we learnt in the class to land on a final binary classification model to use. 
#a. Start by building a model using only the previous time period’s change variables

# 1-0. glm_0, Logistic Regression Model building, threshold=0.5
```{r}
# Load necessary libraries
library(caret)
library(pROC)

# Define a function to print model performance statistics
print_model_statistics <- function(confusion_matrix) {
  overall_stats <- confusion_matrix[["overall"]]
  classwise_stats <- confusion_matrix[["byClass"]]
  
  cat("Accuracy :", round(overall_stats[["Accuracy"]], 3), "\n")
  cat("Precision:", round(classwise_stats[["Precision"]], 3), "\n")
  cat("Recall   :", round(classwise_stats[["Recall"]], 3), "\n")
  
  # Calculate F1 score
  precision <- classwise_stats[["Precision"]]
  recall <- classwise_stats[["Recall"]]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  cat("F1 Score :", round(f1_score, 3), "\n")
}

# Train the logistic regression model
glm_0 <- glm(target_variable ~ ., data = train_data, family = "binomial")

# Make probability predictions on the test set
glm_0_prob_predictions <- predict(glm_0, newdata = test_data, type = "response")

# Define threshold_0
threshold_0 <- 0.5

# Convert probabilities to class labels based on the threshold_0
glm_0_predictions <- ifelse(glm_0_prob_predictions >= threshold_0, 1, 0)

# Prepare actual responses and predictions for evaluation
test_data$target_variable <- as.factor(test_data$target_variable)
glm_0_predictions <- as.factor(glm_0_predictions)
levels(glm_0_predictions) <- levels(test_data$target_variable)

# Evaluate the model and print the confusion matrix
glm_0_confusion_matrix <- confusionMatrix(glm_0_predictions, test_data$target_variable)
print(glm_0_confusion_matrix)

# Print model statistics
cat("Model Statistics for glm_0 with Threshold =", threshold_0, ":\n")
print_model_statistics(glm_0_confusion_matrix)

# ROC and AUC
actual_responses <- as.numeric(as.character(test_data$target_variable)) - 1  # Convert factors to numeric binary (0 and 1)
roc_obj_0 <- roc(actual_responses, glm_0_prob_predictions)
cat("AUC for glm_0:", auc(roc_obj_0), "\n")

# Plot the ROC curve
plot(roc_obj_0, main=paste("ROC Curve for glm_0 (Threshold =", threshold_0, ")"))
```

# 1-1. glm_1, Logistic Regression Model building, threshold=0.1
```{r}
# Load necessary libraries
library(caret)
library(pROC)

# Define a function to print model performance statistics
print_model_statistics <- function(confusion_matrix) {
  overall_stats <- confusion_matrix[["overall"]]
  classwise_stats <- confusion_matrix[["byClass"]]
  
  cat("Accuracy :", round(overall_stats[["Accuracy"]], 3), "\n")
  cat("Precision:", round(classwise_stats[["Precision"]], 3), "\n")
  cat("Recall   :", round(classwise_stats[["Recall"]], 3), "\n")
  
  # Calculate F1 score
  precision <- classwise_stats[["Precision"]]
  recall <- classwise_stats[["Recall"]]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  cat("F1 Score :", round(f1_score, 3), "\n")
}

# Train the logistic regression model
glm_1 <- glm(target_variable ~ ., data = train_data, family = "binomial")

# Make probability predictions on the test set
glm_1_prob_predictions <- predict(glm_1, newdata = test_data, type = "response")

# Define a new threshold
threshold_1 <- 0.1  # changing the threshold to 0.1

# Convert probabilities to class labels based on the new threshold
glm_1_predictions <- ifelse(glm_1_prob_predictions >= threshold_1, 1, 0)

# Prepare actual responses and predictions for evaluation
test_data$target_variable <- as.factor(test_data$target_variable)
glm_1_predictions <- as.factor(glm_1_predictions)
levels(glm_1_predictions) <- levels(test_data$target_variable)

# Evaluate the model and print the confusion matrix
glm_1_confusion_matrix <- confusionMatrix(glm_1_predictions, test_data$target_variable)
print(glm_1_confusion_matrix)

# Print model statistics
cat("Model Statistics for glm_1 with Threshold =", threshold_1, ":\n")
print_model_statistics(glm_1_confusion_matrix)

# ROC and AUC
actual_responses <- as.numeric(as.character(test_data$target_variable)) - 1  # Convert factors to numeric binary (0 and 1)
roc_obj_1 <- roc(actual_responses, glm_1_prob_predictions)
cat("AUC for glm_1:", auc(roc_obj_1), "\n")

# Plot the ROC curve
plot(roc_obj_1, main=paste("ROC Curve for glm_1 (Threshold =", threshold_1, ")"))
```
# 1.2 glm_2, Logistic Regression Model buildind
# Before training, the dataset is resampled using the ovun.sample function from the ROSE package, creating a balanced training set.
```{r}
# Load necessary libraries
library(caret)
library(pROC)
library(ROSE)

# Prepare the training data
train_data$target_variable <- as.factor(train_data$target_variable)

# Balance the dataset using ovun.sample
set.seed(1234)  # for reproducibility
balanced_data <- ovun.sample(target_variable ~ ., data = train_data, method = "both", N = 2000)$data  # Adjust N as needed based on the size of your data

# Train the logistic regression model on the balanced dataset
glm_2 <- glm(target_variable ~ ., data = balanced_data, family = "binomial")

# Make probability predictions on the test set
glm_2_prob_predictions <- predict(glm_2, newdata = test_data, type = "response")

# Define your threshold
threshold_2 <- 0.5  # you might want to adjust this based on your requirements

# Convert probabilities to class labels based on the threshold
glm_2_predictions <- ifelse(glm_2_prob_predictions >= threshold_2, 1, 0)

# Prepare actual responses and predictions for evaluation
test_data$target_variable <- as.factor(test_data$target_variable)
glm_2_predictions <- as.factor(glm_2_predictions)
levels(glm_2_predictions) <- levels(test_data$target_variable)

# Evaluate the model and print the confusion matrix
glm_2_confusion_matrix <- confusionMatrix(glm_2_predictions, test_data$target_variable)
print(glm_2_confusion_matrix)

# Function to print model statistics, assuming it's already defined as in previous snippets
cat("Model Statistics for glm_2 with Threshold =", threshold_2, ":\n")
print_model_statistics(glm_2_confusion_matrix)

# ROC and AUC
actual_responses <- as.numeric(as.character(test_data$target_variable)) - 1  # Convert factors to numeric binary (0 and 1)
roc_obj_2 <- roc(actual_responses, glm_2_prob_predictions)
cat("AUC for glm_2:", auc(roc_obj_2), "\n")

# Plot the ROC curve
plot(roc_obj_2, main=paste("ROC Curve for glm_2 (Threshold =", threshold_2, ")"))


```

# 1.3 glm_3, Logistic Regression Model buildind 
# determine the optimal threshold based on maximizing the sum of sensitivity and specificity.
```{r}
# Load necessary libraries
library(caret)
library(pROC)

# Define a function to print model performance statistics
print_model_statistics <- function(confusion_matrix) {
  overall_stats <- confusion_matrix[["overall"]]
  classwise_stats <- confusion_matrix[["byClass"]]
  
  cat("Accuracy :", round(overall_stats[["Accuracy"]], 3), "\n")
  cat("Precision:", round(classwise_stats[["Precision"]], 3), "\n")
  cat("Recall   :", round(classwise_stats[["Recall"]], 3), "\n")
  
  # Calculate F1 score
  precision <- classwise_stats[["Precision"]]
  recall <- classwise_stats[["Recall"]]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  cat("F1 Score :", round(f1_score, 3), "\n")
}

# Train the logistic regression model
glm_3 <- glm(target_variable ~ ., data = train_data, family = "binomial")

# Make probability predictions on the test set
glm_3_prob_predictions <- predict(glm_3, newdata = test_data, type = "response")

# ROC and AUC
actual_responses <- as.numeric(as.character(test_data$target_variable)) - 1  # Convert factors to numeric binary (0 and 1)
roc_obj_3 <- roc(actual_responses, glm_3_prob_predictions)

# Find the thresholds and the one that maximizes sensitivity + specificity
coords_obj_3 <- coords(roc_obj_3, "best", best.method = "youden", ret = c("threshold", "sensitivity", "specificity"))
optimal_threshold_3 <- coords_obj_3["threshold"]

# Ensure that optimal_threshold_3 is a single numeric value and not a list or any other complex structure
optimal_threshold_3 <- as.numeric(optimal_threshold_3)
cat("Optimal Threshold:", optimal_threshold_3, "\n")

# Simplify the structure of glm_3_prob_predictions to a plain numeric vector without any names or other attributes
glm_3_prob_predictions <- unname(glm_3_prob_predictions)

# Convert probabilities to class labels based on the optimal threshold
glm_3_predictions <- ifelse(glm_3_prob_predictions >= optimal_threshold_3, 1, 0)

# Prepare actual responses and predictions for evaluation
test_data$target_variable <- as.factor(test_data$target_variable)
glm_3_predictions <- as.factor(glm_3_predictions)
levels(glm_3_predictions) <- levels(test_data$target_variable)

# Evaluate the model and print the confusion matrix
glm_3_confusion_matrix <- confusionMatrix(glm_3_predictions, test_data$target_variable)
print(glm_3_confusion_matrix)

# Print model statistics
cat("Model Statistics for glm_3 with Optimal Threshold =", optimal_threshold_3, ":\n")
print_model_statistics(glm_3_confusion_matrix)

cat("AUC for glm_3:", auc(roc_obj_3), "\n")

# Plot the ROC curve
plot(roc_obj_3, main=paste("ROC Curve for glm_3 (Optimal Threshold =", optimal_threshold_3, ")"))


```


## 1.4 glm_4, improved Logistic Regression Model buildind, "Uses SMOTE for balancing, optimizes threshold by maximizing sensitivity + specificity" 
```{r}
# Load necessary libraries
library(caret)
library(pROC)
library(DMwR) # For SMOTE

# Define a function to print model performance statistics
print_model_statistics <- function(confusion_matrix) {
  overall_stats <- confusion_matrix[["overall"]]
  classwise_stats <- confusion_matrix[["byClass"]]
  
  cat("Accuracy :", round(overall_stats[["Accuracy"]], 3), "\n")
  cat("Precision:", round(classwise_stats[["Precision"]], 3), "\n")
  cat("Recall   :", round(classwise_stats[["Recall"]], 3), "\n")
  
  # Calculate F1 score
  precision <- classwise_stats[["Precision"]]
  recall <- classwise_stats[["Recall"]]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  cat("F1 Score :", round(f1_score, 3), "\n")
}

# Balancing the dataset with SMOTE
set.seed(12345)  # For reproducibility of results
balanced_data <- SMOTE(target_variable ~ ., train_data, perc.over = 100, k = 5, perc.under = 200)

# Train the logistic regression model on the balanced data
glm_4 <- glm(target_variable ~ ., data = balanced_data, family = "binomial")

# Make probability predictions on the test set
glm_4_prob_predictions <- predict(glm_4, newdata = test_data, type = "response")

# ROC and AUC
actual_responses <- as.numeric(as.character(test_data$target_variable)) - 1  # Convert factors to numeric binary (0 and 1)
roc_obj_4 <- roc(actual_responses, glm_4_prob_predictions)

# Find the thresholds and the one that maximizes sensitivity + specificity
coords_obj_4 <- coords(roc_obj_4, "best", best.method = "youden", ret = c("threshold", "sensitivity", "specificity"))
optimal_threshold_4 <- coords_obj_4["threshold"]

# Ensure that optimal_threshold_4 is a single numeric value and not a list or any other complex structure
optimal_threshold_4 <- as.numeric(optimal_threshold_4)
cat("Optimal Threshold:", optimal_threshold_4, "\n")

# Simplify the structure of glm_4_prob_predictions to a plain numeric vector without any names or other attributes
glm_4_prob_predictions <- unname(glm_4_prob_predictions)

# Convert probabilities to class labels based on the optimal threshold
glm_4_predictions <- ifelse(glm_4_prob_predictions >= optimal_threshold_4, 1, 0)

# Prepare actual responses and predictions for evaluation
test_data$target_variable <- as.factor(test_data$target_variable)
glm_4_predictions <- as.factor(glm_4_predictions)
levels(glm_4_predictions) <- levels(test_data$target_variable)

# Evaluate the model and print the confusion matrix
glm_4_confusion_matrix <- confusionMatrix(glm_4_predictions, test_data$target_variable)
print(glm_4_confusion_matrix)

# Print model statistics
cat("Model Statistics for glm_4 with Optimal Threshold =", optimal_threshold_4, ":\n")
print_model_statistics(glm_4_confusion_matrix)

cat("AUC for glm_4:", auc(roc_obj_4), "\n")

# Plot the ROC curve
plot(roc_obj_4, main=paste("ROC Curve for glm_4 (Optimal Threshold =", optimal_threshold_4, ")"))

```

```{r}
library(caret)
library(DMwR) # for SMOTE

# Prepare the training data
train_data <- na.omit(train_data)

# Convert character columns to factors
train_data <- data.frame(lapply(train_data, function(x) {
    if(is.character(x)) {
        return(as.factor(x))
    } else {
        return(x)
    }
}))

# Ensure target_variable is a factor
train_data$target_variable <- as.factor(train_data$target_variable)

# Check and print the original distribution of the target variable
print(table(train_data$target_variable))

set.seed(1)
# Perform SMOTE and balance the training data
train_data_balanced <- SMOTE(target_variable ~ ., train_data, perc.over = 100, k = 5) # Adjust perc.over and k as needed

# Check and print the distribution of the target variable after SMOTE
print(table(train_data_balanced$target_variable))

# Train the logistic regression model
glm_4 <- glm(target_variable ~ ., data = train_data_balanced, family = "binomial")

# Make predictions on the test set
glm_4_prob_predictions <- predict(glm_4, newdata = test_data, type = "response")

# Create a ROC curve and calculate AUC
pred_obj <- prediction(glm_4_prob_predictions, test_data$target_variable)
roc_obj_4 <- performance(pred_obj, measure = "tpr", x.measure = "fpr")
auc_obj_4 <- performance(pred_obj, measure = "auc")
auc_glm_4 <- as.numeric(auc_obj_4@y.values)

# Plot ROC curve
plot(roc_obj_4, main = paste("ROC Curve - AUC =", round(auc, 3)), col = "blue", lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "red")

# Find the optimal threshold
perf_obj <- performance(pred_obj, "sens", "spec")
opt_index <- which.max(rowSums(data.frame(sensitivity = unlist(perf_obj@y.values), specificity = unlist(perf_obj@x.values))))
opt_threshold <- perf_obj@alpha.values[[1]][opt_index]
cat("Optimal Threshold:", opt_threshold, "\n")

# Use the optimal threshold to make class predictions
glm_4_predictions <- ifelse(glm_4_prob_predictions >= opt_threshold, 1, 0)
glm_4_predictions <- as.factor(glm_4_predictions)
levels(glm_4_predictions) <- levels(test_data$target_variable)

# Create a confusion matrix
glm_4_confusion_matrix <- confusionMatrix(glm_4_predictions, test_data$target_variable)
print(glm_4_confusion_matrix)

# Extract performance metrics
overall_stats <- glm_4_confusion_matrix[["overall"]]
classwise_stats <- glm_4_confusion_matrix[["byClass"]]

# Calculate and print performance metrics
precision <- classwise_stats[["Precision"]]
recall <- classwise_stats[["Recall"]]
f1_score <- 2 * (precision * recall) / (precision + recall)
cat("Accuracy :", round(overall_stats[["Accuracy"]], 3), "\n")
cat("Precision:", round(precision, 3), "\n")
cat("Recall   :", round(recall, 3), "\n")
cat("F1 Score :", round(f1_score, 3), "\n")
cat("AUC      :", round(auc, 3), "\n")
```

#Performace comparison each model
```{r}
# Extract statistics from confusion matrix and ROC object
get_model_stats <- function(conf_matrix, roc_obj) {
  overall_stats <- conf_matrix[["overall"]]
  class_stats <- conf_matrix[["byClass"]]
  auc_value <- auc(roc_obj)
  
  accuracy <- overall_stats["Accuracy"]
  precision <- class_stats["Precision"]
  recall <- class_stats["Sensitivity"]  # Sensitivity is the same as Recall
  specificity <- class_stats["Specificity"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  sum_sens_spec <- recall + specificity
  
  return(c(accuracy, precision, recall, f1_score, sum_sens_spec, auc_value))
}

# Assuming threshold values are numeric
threshold_0 <- 0.5
threshold_1 <- 0.1
# Extract numeric values from the 'optimal_threshold' if they are not simple numeric values
optimal_threshold_3_value <- optimal_threshold_3  # Extract the actual value if it's in a complex structure
optimal_threshold_4_value <- optimal_threshold_4  # Extract the actual value if it's in a complex structure

# Get statistics for each model
glm_0_stats <- c("glm_0", "threshold=0.5", threshold_0, get_model_stats(glm_0_confusion_matrix, roc_obj_0))
glm_1_stats <- c("glm_1", "threshold=0.1", threshold_1, get_model_stats(glm_1_confusion_matrix, roc_obj_1))
glm_2_stats <- c("glm_2", "Resampled with ROSE", threshold_2, get_model_stats(glm_2_confusion_matrix, roc_obj_2))
glm_3_stats <- c("glm_3", "Maximized Sensitivity and Specificity", optimal_threshold_3_value, get_model_stats(glm_3_confusion_matrix, roc_obj_3))
glm_4_stats <- c("glm_4", "SMOTE, Maximized Sensitivity and Specificity", optimal_threshold_4_value, get_model_stats(glm_4_confusion_matrix, roc_obj_4))

# Combine into a matrix
model_metrics <- rbind(glm_0_stats, glm_1_stats, glm_2_stats, glm_3_stats, glm_4_stats)

colnames(model_metrics) <- c("Model", "Description", "Threshold", "Accuracy", "Precision", "Recall", "F1 Score", "Sum of Sensitivity and Specificity", "AUC")

print(model_metrics)


```


